Supplementary Material for ``Evaluating infectious disease forecasts with allocation scoring rules''

### 2023-11-26 alloscore_manuscript/supplement.rnw
more about quantile functions

We recall how quantiles arise as solutions to a probabilistic decision problem. Let $Y$ be a random variable
representing the future level of an undesirable outcome such as severe COVID incidence. Let $x$ be a decision variable
representing the possible levels of some costly counter-measure, such as procurement of monoclonal antibody treatments,
that can be taken at a cost $C$ per unit in preparation for $Y$.  A decision maker must decide on a level $x$ of investment in the
counter-measure, and wishes to avoid excesses in either the expediture $Cx$ or the shortage $(y-x)_{+}$ when $Y=y$ is
realized. To formalize the trade-off between these potential excesses we quantify the loss associated with a unit of
shortage by a constant $L>C$ (which assumes that the counter-measure has some economic value) and combine the total shortage loss with expenditure into a \emph{loss function}\footnote
{This does involve a confusing use of the word \emph{loss} to refer to two different quantities, but this seems to be
an ingrained and unavoidable habit in the literature.}
\[
l(x,y) = Cx + L(y-x)_{+}.
\] 
The decision problem is then to select a random future loss $l(x,Y)$ in a way that aligns with the preference that $l(x,y)$ be
as low as possible given any realization $Y=y$.

@@

The expected loss is an affine transformation of the convex expected shortage (c.f. \eqref{eqn:convex-esf}). Therefore
$\mathbb{E}_{F}[l(x,Y)]$ is also is convex and has right and left derivatives $D_{\pm}\mathbb{E}_{F}[l(x,Y)]$ at every $x$. 
Because these derivatives exist everywhere, a necessary condition for $x^{\star}$ to
minimize $\mathbb{E}_{F}[l(x,Y)]$ is that $D_{+}\mathbb{E}_{F}[l(x^{\star},Y)] \geq 0$ and 
$D_{-}\mathbb{E}_{F}[l(x^{\star},Y)] \leq 0$, and because of convexity, this condition is also sufficient. From \eqref{eqn:dsfl} 
and \eqref{eqn:dsfr} this means that
\begin{align}
D_{+}\mathbb{E}_{F}[l(x^{\star},Y)] = C + L(F(x^{\star}) - 1) \geq 0 \geq 
D_{-}\mathbb{E}_{F}[l(x^{\star},Y)] = C + L(F(x^{\star}-) - 1) \label{eqn:expQd-ineq}
\end{align}
which rearranges with $\alpha = 1-C/L$ to 
\begin{align}
F(x^{\star}) \geq \alpha \geq F(x^{\star}-). \label{eqn:F-alpha-ineq}
\end{align}
Note that because $F(x)$ and $F(x-)$ are right and left continuous, repectively, the set $\{x \mid F(x) \geq \alpha\}$ is closed
on the left and the set $\{x \mid \alpha \geq F(x-)\}$ is closed on the right.
Therefore, \eqref{eqn:F-alpha-ineq} implies that
\begin{align}
\min\{x \mid F(x) \geq \alpha\}  \leq x^{\star} \leq \max\{x \mid \alpha \geq F(x-)\}. \label{eqn:set-alpha-ineq}
\end{align}
We call $q^{-}_{\alpha, F} := \min\{x \mid F(x) \geq \alpha\}$ and $q^{+}_{\alpha, F} := \max\{x \mid F(x-) \leq \alpha\}$
the left and right quantiles of $F$ (for probability level $\alpha$) and any element $q_{\alpha, F} \in [q^{-}_{\alpha, F}, q^{+}_{\alpha, F}]$
a quantile of $F$.  Thus $x^{\star}$ minimizes the expected loss \eqref{eqn:expQloss} and gives an optimal sultion to the decision problem
if and only if it is a quantile $q_{\alpha, F}$.

Quantiles equivalently arise when the decision problem is defined in terms of the random \emph{oportunity} loss 
\begin{align}
l_o(x,Y) := l(x,Y) - l(Y,Y) = Cx + L(Y-x)_{+} - CY \label{eqn:opp-loss}
\end{align}
which expresses how much more loss is realized by the decision $x$ than an oracle would have incurred, knowing to invest exactly
the future value of $Y$. The optimal decision for $\Ex_F[l_o(x,Y)]$ is the same as for $\Ex_F[l(x,Y)]$ since the term $-C\Ex_F[Y]$ is 
constant in $x$, leading again to the inequalities \eqref{eqn:expQd-ineq}.

@@

@@

But given an auxiliary decision problem $\mathcal{D}_{\mathrm{Aux}} = \{\mathcal{X}, \mathcal{Y}, l(x,y)\}$, such as the
allocation problem from the main text or the cost-loss problem from Section \ref{sec:quantiles_shortage}, we can
produce a scoring rule $S_l$ that is automatically proper using the Bayes act formalism introduced in Section \ref
{sec:methods.detailed.specific_allocation}. This works by defining $S_l(F, y)$ as the value $l(x^F,y)$ already given by  
$\mathcal{D}_{\mathrm{Aux}}$ for the Bayes act $x^F$; that is, the action which \emph{by design} will be assessed via
the forecast $F$ to have the lowest possible expected loss $\Ex[l(x^F,Y^F)]$. In statistical decision theory, 
$\Ex[l(x^F,Y^F)]= \Ex_F[l(x^F,Y)]$ is 
sometimes\footnote{A clear exposition of basic decision theory using this terminology is Chapter 8 of \cite{degroot2005optimal}.
Unfortunately other influential sources such as \cite{berger2013statistical} use ``Bayes risk'' to refer to the expected loss
for a general $x$.} 
refered to as the \emph{Bayes risk} of $F$ relative to $\mathcal{D}_{\mathrm{Aux}}$.

>>>>

We recall how quantiles arise as solutions to a probabilistic decision problem. Let $Y$ be a random variable
representing the future level of an undesirable outcome such as severe COVID incidence. Let $x \in \mathbb{R}$ be a decision variable
representing the possible levels of some costly counter-measure, such as procurement of monoclonal antibody treatments,
that can be taken at a cost $C>0$ per unit in preparation for $Y$.  A decision maker must decide on a level $x$ of investment in the
counter-measure, and wishes to avoid excesses in either the expediture $Cx$ or the shortage $(y-x)_{+}$ when $Y=y$ is
realized. To formalize the trade-off between these potential excesses we quantify the loss associated with a unit of
shortage by a constant $L>C$ (which assumes that the counter-measure has some economic value) and combine the total shortage loss with expenditure into a \emph{loss function}\footnote
{This does involve a confusing use of the word \emph{loss} to refer to two different quantities, but this seems to be
an ingrained and unavoidable habit in the literature.}
\[
l(x,y) = Cx + L(y-x)_{+}.
\] 
The decision problem is then to select a random future loss $l(x,Y)$ in a way that aligns with the preference that $l(x,y)$ be
as low as possible given any realization $Y=y$.

@@

The expected loss is an affine transformation of the convex expected shortage (c.f. \eqref{eqn:convex-esf}). Therefore
$\mathbb{E}_{F}[l(x,Y)]$ is also is convex and has right and left derivatives $D_{\pm}\mathbb{E}_{F}[l(x,Y)]$ at every $x$. 
Because these derivatives exist everywhere, a necessary condition for $x^{\star}$ to
minimize $\mathbb{E}_{F}[l(x,Y)]$ is that $D_{+}\mathbb{E}_{F}[l(x^{\star},Y)] \geq 0$ and 
$D_{-}\mathbb{E}_{F}[l(x^{\star},Y)] \leq 0$, and because of convexity, this condition is also sufficient. From \eqref{eqn:dsfl} 
and \eqref{eqn:dsfr} this means that
\begin{align}
D_{+}\mathbb{E}_{F}[l(x^{\star},Y)] = C + L(F(x^{\star}) - 1) \geq 0 \geq 
D_{-}\mathbb{E}_{F}[l(x^{\star},Y)] = C + L(F(x^{\star}-) - 1) \label{eqn:expQd-ineq}
\end{align}
which rearranges with $\alpha = 1-C/L$ to 
\begin{align}
F(x^{\star}) \geq \alpha \geq F(x^{\star}-). \label{eqn:F-alpha-ineq}
\end{align}
Note that because $F(x)$ and $F(x-)$ are right and left continuous, repectively, and $0<\alpha <1$, the set 
$\{x \mid F(x) \geq \alpha\}$ is closed
on the left and the set $\{x \mid \alpha \geq F(x-)\}$ is closed on the right.
Therefore, \eqref{eqn:F-alpha-ineq} implies that
\begin{align}
\min\{x \mid F(x) \geq \alpha\}  \leq x^{\star} \leq \max\{x \mid \alpha \geq F(x-)\}. \label{eqn:set-alpha-ineq}
\end{align}
We call $q^{-}_{\alpha, F} := \min\{x \mid F(x) \geq \alpha\}$ and $q^{+}_{\alpha, F} := \max\{x \mid F(x-) \leq \alpha\}$
the left and right quantiles of $F$ (for probability level $\alpha$) and any element $q_{\alpha, F} \in [q^{-}_{\alpha, F}, q^{+}_{\alpha, F}]$
a quantile of $F$.  The \emph{quantile function} for $F$, which we write as either $Q_F(\alpha)$ or $F^{-1}(\alpha)$, is the 
\emph{set-valued} function that maps $\alpha \in (0,1)$ to to the set $[q^{-}_{\alpha, F}, q^{+}_{\alpha, F}]$. 
Thus $x^{\star}$ minimizes the expected loss \eqref{eqn:expQloss} and gives an optimal solution to the decision problem
if and only if $x^{\star} \in Q_F(\alpha)$.

\subsection{Quantile functions}
\label{sec:quantile-functions}

For future reference, we record several key properties of quantile functions.

The probability levels $\{\alpha_i\}$ for which $\#Q_F(\alpha_i)>1$ form a discrete subset of $(0,1)$ and correspond to
the non-zero width intervals $[q_{\alpha_i,F}^{-},q_{\alpha_i,F}^{+}]$ where $F$ is constant with values 
$\{\alpha_i\}$.  Conversely, if $F$ is strictly increasing on a (Borel) set $A\subset \mathbb{R}$, then the restriction
$Q_{F}\big|_{F(A)}=F^{-1}\big|_{F(A)}$ of $Q_F$ to $F(A)$ is in fact a real-valued function which is left-continuous
and provides an inverse to $F$ on $F(A)$. If the the support $\mathrm{supp}(F) = \{x \mid 0<F(x)<1\}$ of $F$ is such an
$A$, then extending $F^{-1}$ by $F^{-1}(0)=\inf(\mathrm{supp}(F))$ and $F^{-1}(1)=\sup(\mathrm{supp}(F))$ provides an
inverse to $F$ on $F(\mathbb{R})\subset [0,1]$. If $F$ has a point mass at $x\in A$, so that
$\{\alpha \mid F(x-)< \alpha < F(x)\}$ is a non-empty set disjoint from $F(A)$, then $Q_F$ takes the constant value $x$ 
on the closure $\{\alpha \mid F(x-)\leq \alpha \leq F(x)\}$. Conversely, if $F$ has no discrete component on $A$, 
then $Q_{F}\big|_{F(A)}$ is strictly increasing.

Moreover, it can be said that $Q_F$ is increasing as a set-valued function on $\mathbb{R}$ in the generalized sense that 
$(q_{\alpha, F} - q_{\beta, F})(\alpha-\beta) \geq 0$ whenever
$q_{\alpha, F} \in Q_F(\alpha)$ and $q_{\beta, F} \in Q_F(\beta)$, that is, the set 
$\mathrm{graph}(Q_F) = \{(\alpha, q) \mid q \in Q_F(\alpha)\}$ has no downward sloping secants. Conversely, given such an 
increasing set-valued function $Q$ on $(0,1)$, we can construct a right-continuous function $F_Q$ from $\mathbb{R}$ to $[0,1]$ 
which will be the cdf of the random variable $Y_Q := \min(Q(U))$ where $U \sim \mathrm{Unif}[0,1]$.

\subsection{Opportunity relative to an oracle}
Quantiles equivalently arise when the decision problem is defined in terms of the random \emph{opportunity} loss 
\begin{align}
l_o(x,Y) := l(x,Y) - l(Y,Y) = Cx + L(Y-x)_{+} - CY \label{eqn:opp-loss}
\end{align}
which expresses how much more loss is realized by the decision $x$ than an oracle would have incurred, knowing to invest exactly
the future value of $Y$. The optimal decision for $\Ex_F[l_o(x,Y)]$ is the same as for $\Ex_F[l(x,Y)]$ since the term $-C\Ex_F[Y]$ is 
constant in $x$, leading again to the inequalities \eqref{eqn:expQd-ineq}.

@@

The constraint now determines $\tau$ (and hence the Bayes act) through $\sum_{i=1}^{N} q_{\tau,F_i} = K.$
This equation can be restated as the requirement that 
\begin{align}
K \in TQ_F(\tau) \label{eqn:tau-inclusion}
\end{align}
where the set-valued function
\begin{align}
TQ_F(\tau) := \sum_{i=1}^{N} Q_{F_i}(\tau) = \left[\sum_{i=1}^{N} q^{-}_{\tau,F_i}, \sum_{i=1}^{N} q^{+}_{\tau,F_i}\right]
\end{align}
is defined using interval addition $[a,b] + [c,d] = [a+c, b+d]$.

$TQ_F$ satisfies the conditions mentioned in section \ref{sec:quantile-functions} for being a quantile function, and so there
is a random variable $TY_F$ with cdf $F_{T} := F_{TQ_F}$.  From this perspective, the problem of finding $\tau$ becomes the calculation 
of $F_{T}(K) = \mathbb{P}(TY_F \leq K)$, making clear the existence of a solution $\tau^{\star}$ to \eqref{eqn:tau-inclusion}.

Two awkward features of the quantile representation of the Bayes act can arise.  First, point masses in the $F_i$ create point masses for $F_T$ which may cause $\tau^{\star}$ to not be the unique solution to \eqref{eqn:tau-inclusion}. Secondly, if more than 
one $Q_{F_i}(\tau^{\star})$ is a positive-width interval, then the Bayes act will not be unique in these coordinates, and generically
not all points in the $Q_{F_i}(\tau^{\star})$ will be the coordinate of a Bayes act.

It is important to note that $\lambda$ depends on the forecast $F$ and the constraint level $K$. Thus while $\lambda = L(1-\tau)$ can be
interpreted as a kind of ``cost'' imposed by the constraint in the allocation problem which is analogous to $C = L(1-\alpha)$ in the  
the cost-lost problem of section \ref{sec:quantiles_shortage}, it does not serve to define the allocation loss function in the way that $C$ defines \eqref{eqn:expQloss}. $\lambda$ is rather a parameter that must be found, given the pair $F$ and $K$.

@@

But given an auxiliary decision problem $\mathcal{D}_{\mathrm{Aux}} = \{\mathcal{X}, \mathcal{Y}, l(x,y)\}$, such as the
allocation problem from the main text or the cost-loss problem from Section \ref{sec:quantiles_shortage}, we can
produce a scoring rule $S_l$ that is automatically proper using the Bayes act formalism introduced in Section \ref
{sec:methods.detailed.specific_allocation}. This works by defining $S_l(F, y)$ as the value $l(x^F,y)$ already given by  
$\mathcal{D}_{\mathrm{Aux}}$ for the Bayes act $x^F$; that is, the action which \emph{by design} will be assessed via
the forecast $F$ to have the lowest possible expected loss $\Ex[l(x^F,Y^F)]$. In statistical decision theory, 
$\Ex[l(x^F,Y^F)]= \Ex_F[l(x^F,Y)]$ is 
sometimes\footnote{A clear exposition of basic decision theory using this terminology is Chapter 8 of \cite{degroot2005optimal}.
Unfortunately other influential sources use ``Bayes risk'' to refer to a wider variety of quantities. In 
\cite{berger2013statistical} for example, the Bayes risk can be both the optimal expected loss 
for $F$ (see p. 17) or the expected loss for a general $x$ with respect to $F$ (see p. 6).} 
refered to as the \emph{Bayes risk} of $F$ relative to $\mathcal{D}_{\mathrm{Aux}}$.
